Moisture effects on leaf litter are species-specific and result in non-additive flammability in mixed conifer forest

### 2019-07-26 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Draft sent to Rita for approval

\nwords{7600}
\ntables{X}
\nfig{X}
\nref{X}

@@

  \item Moisture content is a strong determinant of forest fuel flammability
  and varies on small spatial scales and over short time periods. Empirical
  modeling of equilibrium dead fine fuel moisture is well established, but less
  is known about the mechanisms by which species-specific litter traits
  influence litter moisture dynamics, particularly the processes of moisture
  absorption and desorption. Characterizing how species’ leaf litter retain and
  release moisture through time is critical to the assessment of fuel’s
  availability to burn.
  \item Natural forests are often comprised of
  mixed stands, leading to mixtures of leaf litter on the surface fuelling the
  fires. Multiple studies have established the existence of non-additive
  effects in the flammability of dry litter, but we lack an understanding of
  moisture synergies in mixed litter and of moisture effects on the
  flammability of mixed litter.
  \item The litter from eight species of a
  mixed-conifer forest was saturated and allowed to dry to determine moisture
  absorption capacity and desorption rates. Burn trials were performed in moist
  litter beds of single species and of mixtures to establish flammability
  response to the dry-down process.
  \item Moisture dynamics varied across species with
  higher specific leaf area being associated with higher maximum absorption and
  lower litter bulk density being associated with faster drying rates. Species
  that produce more aerated litter beds (such as \emph{Q. kelloggii} and
  \emph{Pinus} species) have faster rates of litter drying.
  \item We examined two
  axes of flammability variation (spread rate and fuel consumption) and found
  that these differing moisture dynamics result in time since wetting having
  strongly differing effects on flammability across the major litter types
  typified by different tree genera. We found that litter mixtures
  exhibited consistent non-additivity in flame spread rate in fuel consumption.
  This non-additivity became more positive with fuel drying.
    \item \emph{Synthesis:} Litter mixtures burn with fire
  behavior more similar to that of the most flammable constituent species and
  this effect increases as fuels dry. statement. Such positive feedback effects
  increases the potential for species to exert community scale effects even
  when relatively rare and that can influence community assembly and the
  evolutionary of niche-constructing traits.
  \end{enumerate}
\end{abstract}

@@

Fuel moisture content is an important, and often driving, determinant of
surface fire behavior. Environmental conditions such as temperature, relative
humidity, and wind drive the moisture content of the fuel available to burn
\citep{Kreye+Varner+etal-2018}. In particular, fine dead fuels respond very
readily to environmental changes \citep{Nelson-2001}. Fuel moisture influences
flammability and fire behavior through effects on ignition, spread rate, flame
height, fuel consumption, and heat release, due to the high specific heat
content of water \citep{Rothermel-1972, Nelson-2011}. Moisture acts as a
heat-sink resulting in less energy available for propagation, and the resulting
water vapor will both cool the flaming front and dilute flammable gases
\citep{Albini-1976, Shafizadeh-1977}. Variation in moisture of living fuels has
strong effects on flammability, but moisture of dead excised litter fuels can
also vary. Moisture of litter fuels varies with climate, with seasons, and with
short term weather within a season. Therefore, the rate of drying and
flammability at any particular moisture content are key parameters that
influence fuel flammability throughout the fire season \citep{Nelson-2001}.

@@

Litter packing (measured as litter-bed density or packing ratio) controls the
flammability of dry litter fuels in seasonally dry climates because higher
litter density reduces flammability, especially on the flame spread rate axis.
Leaf litter density is driven by leaf size and shape
\citep{Fonda+Belanger+etal-1998, Scarff+Westoby-2006, Kane+Varner+etal-2008}.
These litter particle effects on litter bed aeration, in turn, have effects on
litter bed flammability \citep{Ganteaume+Marielle+etal-2011,
  Schwilk+Caprio-2011, Kreye+Varner+etal-2013}. These species-specific effects
are well established, however, how moisture dynamics modulate species specific
flammability is poorly known, and could be driven by by leaf traits, emergent
litter litter traits, or both. Species specific effects on litter flammability
are recorded for the forests of the southeastern US \citep{Nowacki+Abrams-2008}
where species-specific litter drying rates play a role in flammability
feedbacks \citep{Kreye+Varner+etal-2013}. Similarly, in mixed-conifer forest of
the western USA, where the fire regime is characterized by surface fires fueled
by plant litter, overstory tree composition drives variation in local fire
behavior \citep{Schwilk+Caprio-2011} as a result of differential litter
flammability cross tree species and as a result of synergistic interactions in
multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}. These flammability
feedbacks may help explain historical fire frequency patterns, but current
estimates of litter flammability in mixed conifer forests are based largely on
burning dry litter.

@@

Flammability is a multi-dimensional characteristic of fuels
\citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing
different components of fire behavior. In litter driven fire, the two major
axes of variation are described by measurements related to flame spread rate
and measurements related to heat transfer. Flame spread rate is easy to measure
in burning trials, but heat transfer can be more difficult and often proxies
such as duration of heating are used \citep{Magalhaes+Schwilk-2012,
  Varner+Kane+etal-2015}. These different components of fire behavior have
differing ecological effects. Species traits may interact such that the
flammability of multi-species mixtures is not the average of the constituent
species. Such non-additive effects have been shown to be synergistic in dry
litter fuels where the flammability of a mixture is driven by the most
flammable species \citep{VanAltena+Logtestjin+etal-2012,
  Magalhaes+Schwilk-2012}. It is possible that similar non-additive effects may
influence litter moisture content, especially if different leaf traits
influence absorptive capacity and drying rate. There is some evidence for
non-additive flammability in moist fuels with one study finding increasing
non-additivity with moisture \citep{Blauw+Wensink+etal-2015}.

We investigated species specific litter moisture dynamics across eight tree
species in a temperate fire-prone forest in California. We determined the
maximum water retention and rate of moisture loss and examined how these vary
in relation to two leaf and litter bed traits. We then examined the influence
of moisture on the flammability across these species. Experimental burn trials
of leaf litter at various moisture levels determined how time since wetting
influenced two axes of flammability characterized by flame spread rate and
total fuel consumed. Previous work described the flammability of these species
in monocultures and in mixtures for oven-dry litter
\citep{Magalhaes+Schwilk-2012} and found consistent positive non-additivity of
flammability where fire behavior of mixtures was driven by the most flammable
species. Here, we examined if this positive non additivity was maintained in
mixtures across varying levels of fuel moisture. The burn trials were also
performed with multi-species litter mixtures to evaluate if species contribute
equally to litter moisture dynamics or if there are non-additive effects as
observed in oven-dried litter. We tested if the non-additive effects on
reported in previous studies occur under ecologically-relevant moisture levels
and whether fuel moisture effects strengthen or mask such non-additivity.

@@

Field sites for this study were located in Sequoia and Kings Canyon National
Parks, California, USA (36 36N, 118 42 W) between 1600 and 2400 m elevation in
mixed-conifer forest. We chose eight dominant tree species that are
representative of this type of forest: \emph{Pinus jeffreyi} Grev. \& Balf.,
\emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl. ex Laws.,
\emph{Abies concolor} (Gord. \& Glend.) Lindl. ex Hildebr., \emph{Abies
  magnifica} A. Murr., \emph{Calocedrus decurrens} (Torr.) Florin,
\emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J.
Buchholz. In this study, we consider litter the top layer of leaves and small
twigs less than 0.625 cm diameter (= 1-hour fuel) that have fallen in the
previous year (mostly undecomposed). Leaf litter was collected in the summer
(mid-June to mid-July) in 2011 and in 2012 from 21 separate sites across the
study area. We collected from a minimum of four populations per species with
each population 9--32\,km distant from its neighbors. An exception to this was
\emph{A. magnifica} which is restricted to the higher elevations and could only
be collected from one area, due to designated wilderness boundaries and
logistical constraints. Within that area, \emph{A. magnifica} litter was
collected from over 10 different individuals. For the remaining species, the
collection involved 2--4 individual trees at each site, separated from one
another by at least 10\,m. The litter was collected 2\,m away from the trunk to
obtain a more uniform sample of leaf and fine branchlet litter, because bark
and heavier branches tend to fall closer to the tree trunk. All litter was air
dried to $< 5$\% fuel moisture by dry weight for weighing and dividing into
moisture trials. The moisture manipulations and subsequent burning trials were
conducted in April--August 2015.

@@

We produced dry-down curves to determine maximum water retainability and the
rate of moisture content loss for each species and for four distinct
three-species mixtures. These curves allowed us to determine the rate at which
a species loses moisture, approximated as exponential decay, and to determine
maximum water content following saturation and gravity draining. Samples from
the eight species (six replicates per species) and four mixtures (5 replicates
per mixture type) were placed in baskets approximately 45 x 45\,cm by 15\,cm
tall constructed of 1\,cm aluminum screen. Samples were 450\,g dry weight
litter per trial and litter depth was standardized to 10\,cm by changing the
horizontal dimensions of the basket. Weight was measured by using a balance
sensitive to 0.1\,g (model XS16001L, Mettler Toledo, Columbus, OH). Samples
were immersed in plastic storage bins filled with water for 24 hours to fully
saturate them. After saturation, the baskets were removed from the water,
allowed to drain for 3 minutes, and samples weighed to measure maximum moisture
content. The samples were then allowed to dry in a controlled environment kept
at 21\,C temperature and 30\% relative humidity, assessed via an iButton sensor
(Maxim Integrated) placed over the samples. Litter baskets were arranged in a
completely randomized design. During the dry-down process three subsamples were
taken at 24 hour intervals and weighed. These subsamples were then oven dried
for 24 hours at 100\,C and re-weighed to assess moisture content. This was
repeated until the final sample from a basket was less than 10\% moisture.

With only eight species in this study, our power to test leaf trait effects on
the observed moisture dynamics was low, so we decided \emph{a priori} to test
the effect of one leaf trait likely to influence moisture absorption, specific
leaf area, and one litter trait known to directly effect flammability and
likely to influence drying rate, litter bulk density. Specific leaf area was
taken from leaf freshly fallen litter leaf area and leaf dry mass from 5 individuals per
species. This measurement included petiole and for the scale-leaved species included the full excised branchlet, so this is not directly comparable to typical specific leaf area measurements on fresh leaves. Litter bulk density was obtained in this
experiment for each basket by dividing the dry mass (450\,g) by the estimated
litter volume (basket width x length x litter depth).

@@

We performed flammability tests across multiple moisture levels on monospecific
litter beds and on mixed litter beds with each species or mixture replicated 5
times at each moisture level. We used the initial dry-down experiment to guide
our timing so as to attempt to burn samples over a range of moisture contents.
We aimed for moisture contents from 10\% to 100\% on a dry mass basis as
preliminary trials on damp fuels suggested that across these fuels ignition is
unlikely above 100\% fuel moisture. To expand our data on the dry end of this
gradient, we supplemented these data with data from an earlier experiment on
oven-dried litter \citep{Magalhaes+Schwilk-2012}. This supplemental data added
five replicates per species at an oven dried moisture content of $<$ 5\%. We
assigned these replicates a time since wetting value of 144 hours equivalent to
the longest time since drying for the wetted samples. Flammability was assessed
using a 150\,cm long burn table, 15\,cm wide and 15\,cm tall, in which leaf
litter was placed and the table gently shaken to allow settling. Burning trials
all used 450\,g (dry weight) of litter. For three-species mixtures, this was
150\,g per species. This design allows fire to reach constant flame spread
rates \citep{Magalhaes+Schwilk-2012}. Two graduated metal rulers equally spaced
along the apparatus allowed visual determination of maximum flame height.
Litter was ignited with a propane torch; maximum adiabatic flame temperature of
1899 C, \url{http://www.benzomatic.com/products/fuel.aspx}) at one end of the
apparatus and allowed to burn until extinction. If a sample failed to ignite
after 30 seconds, no further attempt was made and if three samples of a species
at a particular moisture level failed to ignite, no further attempts were made
on that species/moisture combination and the replicate was allowed to continue
drying. A timer was used to record total time to ignition and duration of
flaming combustion. The flammability trials followed the methodology detailed
in de Magalhães and Schwilk \citep{Magalhaes+Schwilk-2012}.

@@

Where $m$ is moisture content at time $t$, and $\lambda$ is the exponential
decay coefficient. To estimate these coefficients we fit a linear mixed-effects
model by first taking the natural logarithm of moisture content, with time
since wetting and species as fixed effects nested within replicate basket. We
extracted the coefficients of the fitted model to describe maximum water
retainability ($m_{max}$; g of water per g of dry mass) and desiccation rate
($\lambda$; hour$^{-1}$) for each species. All analyses were conducted in R
\citep{RCoreTeam-2019}. We fit models with R packages `lme4'
\citep{Bates_Machler_etal-2015} and `afex' \citep{Singmann_Bolker_etal-2017}.
Approximate degrees of freedom and p-values were calculated by the
Kenward-Roger approximation \citep{Kenward_Roger-1997} which is recommended by
Luke \citep{Luke-2017} as producing acceptable Type 1 error rates. We tested if
one leaf trait (specific leaf area) and one litter trait (litter bulk density)
influenced moisture dynamics by fitting linear models with species means as
observations. For each response variable we fit a single linear model with two
predictors and an interaction term.

Our past work \citep{Magalhaes+Schwilk-2012} demonstrated that dry litter
flammability across these eight species was driven by leaf traits and litter
packing. These previously reported flammability trials demonstrated that
species within a genus tended to behave similarly and the two scale-leaved
genera in the Cupressaceae, \emph{Calocedrus} and \emph{Sequoiadendron}, also
behaved similarly to one another. Therefore, to preserve power used taxonomic
groups (\emph{Abies}, Cupressaceae, \emph{Quercus}, and \emph{Pinus}) as our
flammability functional groups and treated these as fixed effects in our
models. The effect of fuel moisture and time since wetting on flammability
parameters were modeled using linear models with fuel moisture or time since
wetting and taxon as fixed effects and average vapor pressure deficit in the
burning room during that trial as a nuisance covariate. We used linear mixed
models to predict spread rate, but fuel consumption tends to exhibit threshold
effects with few intermediate values. Therefore, we used mixed generalized
linear models with a beta regression link with the `glmmADMB` package
\citep{Skaug_Fournier_etal-2016} to model percent fuel consumed as a function
moisture content and of time since wetting.

We investigated possible non-additive effects on moisture and on flammability
by testing if moisture content and flammability of mixtures were predicted by
the average of the constituent species. To choose mixtures, we first grouped
the species according to their maximum water retainability and dessication rate
according to estimated marginal means by species using the R package `emmeans'
\citep{Lenth-2019}. This resulted in four groups of species and we then selected
one representative species from each litter type group and produced all four
possible three-species litter mixtures using those four species. For each
mixture and time since wetting, we predicted moisture content based on the mean
content of the three constituent species at that time since wetting. We
predicted flammability parameters based on the average flammability measures
from individual species trials at each time since wetting and then calculated a
predicted spread rate and fuel consumption as the average of these three values
for the mixture. The expectation from the null model was that the difference
between observed and predicted values of flammability will be zero. We tested
if there were significant departures from the null (zero) with a mixed effects
linear model. For flammability parameters, we fit linear mixed effect models
with the observed-predicted flammability value as the dependent variable and
used Wald tests to test if the centered intercept was different than zero (a
measure of overall non-additivity) and if the slope with time since wetting was
different from zero (indicating non-additivity changing with drying).

@@

There were marked differences in the extent of moisture species retain
initially following wetting and draining by gravity (all $P$ < 0.0001, Fig.
\ref{fig:drydown} and Table \ref{tab:drydown}). \emph{Q. kelloggii}, the only
broadleaf, is the species that retained the most moisture initially.
\emph{Quercus}, like the pines, had a rapid drying rate (Fig. \ref{fig:drydown}
and SI Fig. 1). Calculating estimated marginal means following the linear
mixed-effects model discriminated the species into three different maximum
water retention groups: 1) \emph{Q. kelloggii}; 2) \emph{P. lambertiana} and
\emph{C. decurrens}; and 3) \emph{A. magnifica}, \emph{S. giganteum}, \emph{A.
  concolor}, \emph{P. jeffreyi} and \emph{P. ponderosa} (SI Fig. 1). There are
also differences in desiccation rates across the eight species and estimated
mixed linear model slopes discriminated species into three groups from lowest
to most rapid: 1) \emph{C.decurrens}, 2) \emph{Abies spp} and \emph{S.
  giganteum}, and 3) \emph{Q. kelloggii} and \emph{Pinus spp} (SI Fig. 2).

\begin{figure}[h]
  \centering
\includegraphics[width=8cm]{figs_tables/fig1_drydown-curves.pdf}
\caption{Dry down curves for eight litter types. Lines show best-fit
  exponential decay curves fit to moisture content on a dry mass basis. Litter
  samples we saturated by immersion for 24 hours and then allowed to drain by
  gravity for 3 min before initial weighing. Colors indicate genera and there
  were significant differences in both maximum water retention and in rate of
  drying (based linear model on log of water content). A small amount of
  horizontal jitter (0.5 hr) was added to this plot to aid in visualizing
  because, although collections were staggered, there were sometimes multiple
  trays collected at a single time point.}
 \label{fig:drydown}
\end{figure}

\begin{table}[h]
  \caption{Linear mixed model results for moisture content as a function of
    time since wetting. Approximate degrees of freedom (df), pseudo F
    statistics and p-values were calculated by the Kenward-Roger approximation.
    \citep{Kenward_Roger-1997}.}
  \label{tab:drydown}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{llll}
  \toprule
Effect & df & $F$ & $P$ \\ 
  \midrule
  hour & 1, 183.73 & 1599.59 & $<$.0001 \\ 
  species & 7, 89.73 & 36.15 & $<$.0001 \\ 
  hour:species & 7, 184.15 & 23.06 & $<$.0001 \\ 
   \bottomrule
\end{tabular}
\end{table}

>>>>

\nwords{7600}
\ntables{4}
\nfig{8}
\nref{39}

@@

      \item Moisture content is a strong determinant of forest fuel
    flammability and varies on small spatial scales and over short time
    periods. Empirical modeling of equilibrium dead fine fuel moisture is well
    established, but less is known about the mechanisms by which
    species-specific litter traits influence litter moisture dynamics.
    Characterizing how species’ leaf litter retain and release moisture through
    time is critical to the assessment of fuel’s availability to burn.
      \item Natural forests are often comprised of mixed stands, leading to
    mixtures of leaf litter on the surface fuelling the fires. Multiple studies
    have established the existence of non-additive effects in the flammability
    of dry litter, but we lack an understanding of moisture synergies in mixed
    litter and of moisture effects on the flammability of mixed litter.
      \item The litter from eight species of a mixed-conifer forest was
    saturated and allowed to dry to determine moisture absorption capacity and
    drying rates. Burn trials were performed in moist litter beds of single
    species and of mixtures to establish flammability response to the dry-down
    process.
      \item Moisture dynamics varied across species with higher specific leaf
    area being associated with higher maximum absorption and lower litter bulk
    density being associated with faster drying rates. Species that produce
    more aerated litter beds (such as \emph{Q. kelloggii} and \emph{Pinus}
    species) have faster rates of litter drying.
      \item We examined two axes of flammability variation (spread rate and
    fuel consumption) and found that these differing moisture dynamics result
    in time since wetting having strongly differing effects on flammability
    across the major litter types typified by different tree taxa.
      \item We found that litter mixtures exhibited consistent non-additivity
    in flame spread rate in fuel consumption. This non-additivity became more
    positive with fuel drying. Litter mixtures burn with fire behavior more
    similar to that of the most flammable constituent species and this effect
    increases as fuels dry.
      \item \emph{Synthesis:} In this temperate mixed conifer forest, the shade
    intolerant species are favored by the increased fire to which their fuels
    contribute Such positive feedback effects increases the potential for
    species to exert community scale effects even when relatively rare and that
    can influence community assembly and the evolutionary of niche-constructing
    traits.
  \end{enumerate}
\end{abstract}

@@

Fuel moisture content is an important, and often driving, determinant of
surface fire behavior. Environmental conditions such as temperature, relative
humidity, and wind drive the moisture content of the fuel available to burn
\citep{Kreye+Varner+etal-2018}. In particular, fine dead fuels respond very
readily to environmental changes \citep{Nelson-2001}. Fuel moisture influences
flammability and fire behavior through effects on ignition, spread rate, flame
height, fuel consumption, and heat release, due to the high specific heat
content of water \citep{Rothermel-1972, Nelson-2001}. Moisture acts as a
heat-sink resulting in less energy available for propagation, and the resulting
water vapor will both cool the flaming front and dilute flammable gases
\citep{Albini-1976, Shafizadeh-1977}. Variation in moisture of living fuels has
strong effects on flammability, but moisture of dead excised litter fuels can
also vary. Moisture of litter fuels varies with climate, with seasons, and with
short term weather within a season. Therefore, the rate of drying and
flammability at any particular moisture content are key parameters that
influence fuel flammability throughout the fire season \citep{Nelson-2001}.

@@

Litter packing (measured as litter-bed density or packing ratio) controls the
flammability of dry litter fuels in seasonally dry climates because higher
litter density reduces flammability, especially on the flame spread rate axis.
Leaf litter density is driven by leaf size and shape
\citep{Fonda+Belanger+etal-1998, Scarff+Westoby-2006, Kane+Varner+etal-2008}.
These litter particle effects on litter bed aeration, in turn, have effects on
litter bed flammability \citep{Ganteaume+Marielle+etal-2011,
  Schwilk+Caprio-2011, Kreye+Varner+etal-2013}. These species-specific effects
are well established. However, how moisture dynamics modulate species specific
flammability is poorly known, and could be driven by by leaf traits, emergent
litter litter traits, or both. Species specific effects on litter flammability
are recorded for the forests of the southeastern US \citep{Nowacki+Abrams-2008}
where species-specific litter drying rates play a role in flammability
feedbacks \citep{Kreye+Varner+etal-2013}. Similarly, in mixed-conifer forest of
the western USA, where the fire regime is characterized by surface fires fueled
by plant litter, overstory tree composition drives variation in local fire
behavior \citep{Schwilk+Caprio-2011} as a result of differential litter
flammability across tree species and as a result of synergistic interactions in
multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}. These flammability
feedbacks may help explain historical fire frequency patterns, but current
estimates of litter flammability in mixed conifer forests are based largely on
burning dry litter.

@@

Flammability is a multi-dimensional characteristic of fuels
\citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing
different components of fire behavior. In litter driven fire, the two major
axes of variation are described by measurements related to flame spread rate
and measurements related to heat transfer. Flame spread rate is easy to measure
in burning trials, but heat transfer can be more difficult and often proxies
such as duration of heating are used \citep{Magalhaes+Schwilk-2012,
  Varner+Kane+etal-2015}. These different components of fire behavior have
differing ecological effects. Species traits may interact such that the
flammability of multi-species mixtures is not the average of the constituent
species. Such non-additive effects are synergistic in dry litter fuels where
the flammability of a mixture is driven by the most flammable species
\citep{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}. It is possible
that similar non-additive effects may influence litter moisture content,
especially if different leaf traits influence absorptive capacity and drying
rate. There is some evidence for non-additive flammability in moist fuels with
one study finding increasing non-additivity with moisture
\citep{Blauw+Wensink+etal-2015}.

We investigated species specific litter moisture dynamics across eight tree
species in a temperate fire-prone forest in California. We determined the
maximum water retention and rate of moisture loss and examined how these vary
in relation to two leaf and litter bed traits. We then examined the influence
of moisture on the flammability across these species. Experimental burn trials
of leaf litter at various moisture levels determined how time since wetting
influenced two axes of flammability characterized by flame spread rate and
total fuel consumed. Previous work described the flammability of these species
in monocultures and in mixtures for oven-dry litter
\citep{Magalhaes+Schwilk-2012} and found consistent positive non-additivity of
flammability where fire behavior of mixtures was driven by the most flammable
species. Here, we examined if this positive non-additivity was maintained in
mixtures across varying levels of fuel moisture. The burn trials were also
performed with multi-species litter mixtures to evaluate if species contribute
equally to litter moisture dynamics or if there are non-additive effects as
observed in oven-dried litter. We tested if the non-additive effects on
reported in previous studies occur under ecologically-relevant moisture levels
and whether fuel moisture effects strengthen or moderate such non-additivity.

@@

Field sites for this study were located in Sequoia and Kings Canyon National
Parks, California, USA (36 36N, 118 42 W) between 1600 and 2400 m elevation in
mixed-conifer forest. We chose eight dominant tree species that are
representative of this type of forest: \emph{Pinus jeffreyi} Grev. \& Balf.,
\emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl. ex Laws.,
\emph{Abies concolor} (Gord. \& Glend.) Lindl. ex Hildebr., \emph{Abies
  magnifica} A. Murr., \emph{Calocedrus decurrens} (Torr.) Florin,
\emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J.
Buchholz. In this study, we consider litter the top layer of leaves and small
twigs less than 0.625 cm diameter (= 1-hour fuel) that have fallen in the
previous year (mostly undecomposed). Leaf litter was collected in the summer
(mid-June to mid-July) in 2012 from 21 separate sites across the
study area. We collected from a minimum of four populations per species with
each population 9--32\,km distant from its neighbors. An exception to this was
\emph{A. magnifica} which is restricted to the higher elevations and could only
be collected from one area, due to designated wilderness boundaries and
logistical constraints. Within that area, \emph{A. magnifica} litter was
collected from over 10 different individuals. For the remaining species, the
collection involved 2--4 individual trees at each site, separated from one
another by at least 10\,m. The litter was collected 2\,m away from the trunk to
obtain a more uniform sample of leaf and fine branchlet litter, because bark
and heavier branches tend to fall closer to the tree trunk. All litter was air
dried to $< 5$\% fuel moisture by dry weight for weighing and dividing into
moisture trials. The moisture manipulations and subsequent burning trials were
conducted in April--August 2015.

@@

We produced dry-down curves to determine maximum water retainability and the
rate of moisture content loss for each species and for four distinct
three-species mixtures. These curves allowed us to determine the rate at which
a species loses moisture, approximated as exponential decay, and to determine
maximum water content following saturation and gravity draining. Samples from
the eight species (six replicates per species) and four mixtures (5 replicates
per mixture type) were placed in baskets approximately 45 x 45\,cm by 15\,cm
tall constructed of 1\,cm aluminum screen. Samples were 450\,g dry weight
litter per trial and litter depth was standardized to 10\,cm by changing the
horizontal dimensions of the basket. Weight was measured by using a balance
sensitive to 0.1\,g (model XS16001L, Mettler Toledo, Columbus, OH). Samples
were immersed in plastic storage bins filled with water for 24 hours to fully
saturate them. After saturation, the baskets were removed from the water,
allowed to drain for 3 minutes, and samples weighed to measure maximum moisture
content. The samples were then allowed to dry in a controlled environment kept
at 21\,C temperature and 30\% relative humidity, assessed via an iButton sensor
(Maxim Integrated) placed over the samples. Litter baskets were arranged in a
completely randomized design. During the dry-down process three subsamples of
5--10\,g were taken at 24 hour intervals and weighed. These subsamples were
then oven dried for 24 hours at 100\,C and re-weighed to assess moisture
content. This was repeated until the final sample from a basket was less than
10\% moisture.

With only eight species in this study, our power to test leaf trait effects on
the observed moisture dynamics was low, so we decided \emph{a priori} to test
the effect of one leaf trait likely to influence moisture absorption, specific
leaf area, and one litter trait known to directly effect flammability and
likely to influence drying rate, litter bulk density. Specific leaf area was
measured on freshly fallen litter by sampling 5 individuals per species. This
measurement included petiole and for the scale-leaved species included the full
excised branchlet. Therefore, this is not directly comparable to typical specific leaf
area measurements on fresh leaves. Litter bulk density was obtained in this
experiment for each basket by dividing the dry mass (450\,g) by the estimated
litter volume (basket width x length x litter depth).

@@

We performed flammability tests across multiple moisture levels on monospecific
litter beds and on mixed litter beds with each species or mixture replicated 5
times at each moisture level. We used the initial dry-down experiment to guide
our timing so as to attempt to burn samples over a range of moisture contents.
We aimed for moisture contents from 10\% to 100\% on a dry mass basis as
preliminary trials on damp fuels suggested that across these fuels ignition
will not occur above 100\% fuel moisture. To expand our data on the dry end of
this gradient, we supplemented these data with data from an earlier experiment
on oven-dried litter \citep{Magalhaes+Schwilk-2012}. This supplemental data
added five replicates per species at an oven dried moisture content of $<$ 5\%.
We assigned these replicates a time since wetting value of 144 hours equivalent
to the longest time since drying for the wetted samples. Flammability was
assessed using a 150\,cm long burn table, 15\,cm wide and 15\,cm tall, in which
leaf litter was placed and the table gently shaken to allow settling. Burning
trials all used 450\,g (dry weight) of litter. For three-species mixtures, this
was 150\,g per species. This design allows fire to reach constant flame spread
rates \citep{Magalhaes+Schwilk-2012}. Two graduated metal rulers equally spaced
along the apparatus allowed visual determination of maximum flame height.
Litter was ignited with a propane torch; maximum adiabatic flame temperature of
1899 C, \url{http://www.benzomatic.com/products/fuel.aspx}) at one end of the
apparatus and allowed to burn until extinction. If a sample failed to ignite
after 30 seconds, no further attempt was made and if three samples of a species
at a particular moisture level failed to ignite, no further attempts were made
on that species/moisture combination and the replicate was allowed to continue
drying. A timer was used to record total time to ignition and duration of
flaming combustion. The flammability trials followed the methodology detailed
in de Magalhães and Schwilk \citep{Magalhaes+Schwilk-2012}.

@@

Where $m$ is moisture content at time $t$, and $\lambda$ is the exponential
decay coefficient. To estimate these coefficients we fit a linear mixed-effects
model by first taking the natural logarithm of moisture content, with time
since wetting and species as fixed effects nested within replicate basket. We
extracted the coefficients of the fitted model to describe maximum water
retainability ($m_{max}$; g of water per g of dry mass) and desiccation rate
($\lambda$; hour$^{-1}$) for each species. All analyses were conducted in R
\citep{RCoreTeam-2019}. We fit models with R packages `lme4'
\citep{Bates_Machler_etal-2015} and `afex' \citep{Singmann_Bolker_etal-2017}.
Approximate degrees of freedom and p-values were calculated by the
Kenward-Roger approximation \citep{Kenward_Roger-1997} which is recommended by
Luke \citeyear{Luke-2017} as producing acceptable Type 1 error rates. We tested if
one leaf trait (specific leaf area) and one litter trait (litter bulk density)
influenced moisture dynamics by fitting linear models with species means as
observations. For each response variable we fit a single linear model with two
predictors and an interaction term.

Our past work \citep{Magalhaes+Schwilk-2012} demonstrated that dry litter
flammability across these eight species was driven by leaf traits and litter
packing. These previously reported flammability trials demonstrated that
species within a genus tended to behave similarly and the two scale-leaved
genera in the Cupressaceae, \emph{Calocedrus} and \emph{Sequoiadendron}, also
behaved similarly to one another. Therefore, to preserve power we used taxonomic
groups (\emph{Abies}, Cupressaceae, \emph{Quercus}, and \emph{Pinus}) as our
flammability functional groups and treated these as fixed effects in our
models. The effect of fuel moisture and time since wetting on flammability
parameters were modeled using linear models with fuel moisture or time since
wetting and taxon as fixed effects and average vapor pressure deficit in the
burning room during that trial as a nuisance covariate. We used linear mixed
models to predict spread rate, but fuel consumption tends to exhibit threshold
effects with few intermediate values. Therefore, we used mixed generalized
linear models with a beta regression link with the `glmmADMB` package
\citep{Skaug_Fournier_etal-2016} to model percent fuel consumed as a function of
moisture content and of time since wetting.

We investigated possible non-additive effects on moisture and on flammability
by testing if moisture content and flammability of mixtures were predicted by
the average of the constituent species. To choose mixtures, we first grouped
the species according to their maximum water retention and drying rate
according to estimated marginal means by species using the R package `emmeans'
\citep{Lenth-2019}. This resulted in four groups of species and we then selected
one representative species from each litter type group and produced all four
possible three-species litter mixtures using those four species. For each
mixture and time since wetting, we predicted moisture content based on the mean
content of the three constituent species at that time since wetting. We
predicted flammability parameters based on the average flammability measures
from individual species trials at each time since wetting and then calculated a
predicted spread rate and fuel consumption as the average of these three values
for the mixture. The expectation from the null model was that the difference
between observed and predicted values of flammability will be zero. We tested
if there were significant departures from the null (zero) with a mixed effects
linear model. For flammability parameters, we fit linear mixed effect models
with the observed-predicted flammability value as the dependent variable and
used Wald tests to test if the centered intercept was different than zero (a
measure of overall non-additivity) and if the slope with time since wetting was
different from zero (indicating non-additivity changing with drying).

@@

There were marked differences in the extent of moisture species retain
initially following wetting and draining by gravity (all $P$ < 0.0001, Fig.
\ref{fig:drydown} and Table \ref{tab:drydown}). \emph{Q. kelloggii}, the only
broadleaf, is the species that retained the most moisture initially.
\emph{Quercus}, like the pines, had a rapid drying rate (Fig. \ref{fig:drydown}
and SI Fig. 1). Calculating estimated marginal means following the linear
mixed-effects model discriminated the species into three different maximum
water retention groups: 1) \emph{Q. kelloggii}; 2) \emph{P. lambertiana} and
\emph{C. decurrens}; and 3) \emph{A. magnifica}, \emph{S. giganteum}, \emph{A.
  concolor}, \emph{P. jeffreyi} and \emph{P. ponderosa} (SI Fig. 1). There are
also differences in drying rates across the eight species and estimated
mixed linear model slopes discriminated species into three groups from lowest
to most rapid: 1) \emph{C. decurrens}, 2) \emph{Abies spp} and \emph{S.
  giganteum}, and 3) \emph{Q. kelloggii} and \emph{Pinus spp} (SI Fig. 2).

\begin{figure}[h]
  \centering
\includegraphics[width=8cm]{figs_tables/fig1_drydown-curves.pdf}
\caption{Dry down curves for eight litter types. Lines show best-fit
  exponential decay curves fit to moisture content on a dry mass basis. Litter
  samples we saturated by immersion for 24 hours and then allowed to drain by
  gravity for 3 min before initial weighing. Colors indicate genera and there
  were significant differences in both maximum water retention and in rate of
  drying (based linear model on log of water content). A small amount of
  horizontal jitter (0.5 hr) was added to this plot to aid in visualizing
  because, although collections were staggered, there were sometimes multiple
  samples collected at a single time point.}
 \label{fig:drydown}
\end{figure}

\begin{table}[h]
  \caption{Linear mixed model results for moisture content as a function of
    time since wetting and species. Approximate degrees of freedom (df), pseudo F
    statistics and p-values were calculated by the Kenward-Roger approximation.
    \citep{Kenward_Roger-1997}. Model coefficients reported in SI Table 1.}
  \label{tab:drydown}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{llll}
  \toprule
Effect & df & $F$ & $P$ \\ 
  \midrule
  hour & 1, 183.73 & 1599.59 & $<$.0001 \\ 
  species & 7, 89.73 & 36.15 & $<$.0001 \\ 
  hour:species & 7, 184.15 & 23.06 & $<$.0001 \\ 
   \bottomrule
\end{tabular}
\end{table}
